\thetitle

### 2020-11-06 writing/mutate.tex
mutate: Revise description of mutators and interesting scenarios

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow
~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
A mutation is a local syntactic change of the code of a program that may
produce a bug.
\ll{Need citations for mutation}
Usually, the outcome of mutating a program
once, dubbed a mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Thus a big part of research on mutation analysis goes into designing
operations that perform mutations, dubbed mutators, that create mutants
that are hard to ``kill.'' Unlike the standard mutators, which \citet{lksfd-popl-2020}
use, too, the mutators we need to evaluate the rational programmer have to be
fine-tuned to generate shallow errors. Typed Racket's type system detects
typos not logical mistakes, and mutators suitable for our purposes
should target the former rather than the latter.

The table in figure~\ref{table:mutation-ops} summarizes our mutators.
For each one, the table provides a short description and an example
mutation it performs. Each mutator is inspired by the more-than-a-decade
long experience of the authors making type-level mistakes in Typed Racket.
We find that these mistakes often take non-trivial effort to debug.

@@

\subsection{Do these mutators give us interesting debugging scenarios?}
\begin{itemize}
\item do these mutators work?
\begin{itemize}
\item now we need to see if these mutators actually produce interesting errors
\begin{itemize}
\item define interesting errors: type error + dynamic err w/ 3 modules on stack in a mixed configuration
\item this section will show if they do this on our benchmarks
\end{itemize}
\item do they actually produce interesting errors? -- for this need plot that's inverted version dyn-err-heatmap + total of mutants per benchmark
\item do we need all of the mutators?
\begin{itemize}
\item each one produces type errors + interesting scenarios
\item we conjecture they represent different kinds of mistakes
\end{itemize}
\end{itemize}
\end{itemize}

Having selected a set of mutators based on our experience, we next need to see if the mutators do in fact produce interesting debugging scenarios in our programs.
We define an interesting debugging scenario as a mutation that
\begin{enumerate}
\item can be caught by the type checker if the program is fully typed,
\item otherwise produces a dynamic error under Erasure, and
\item the stacktrace of the dynamic error contains at least three distinct modules from the program.
\end{enumerate}
The first criterion reflects the fact that a gradual type system can only catch bugs that would have been caught by a type checker.
The second criterion ensures that the bug is detectable by all three semantics.
It is based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in Erasure implies a dynamic type error in both Transient and Natural.
The final criterion conservatively approximates scenarios with the opportunity for interactions between typed and untyped modules. % I say opportunity because it depends on the config
In particular, it takes at least two modules to have typed/untyped module interactions (one of each kind), and having two modules on the stack ensures that the error arose in the interaction of at least those two modules.
The third module comes in because every one of our programs comes with a driver module which is usually uninteresting but should be present in every stacktrace.
With this definition in hand, the rest of this section will analyze how well our mutators produce interesting debugging scenarios.

Beginning with the first criterion of the definition, we analyze how well our mutators produce type errors.

%% Cut from above
An effective set of mutators for type errors should be able to generate
a large number of ill-typed programs. Hence, to validate that our experience
with Typed Racket translates to effective mutators, we analyze the
 mutants of our ten benchmarks. Figure~\ref{fig:mutators-static-hit} shows
 the results. For each benchmark, our mutators generate between 156 and
 18165 ill-typed mutants. Most importantly the collective success rate of
 the mutators  per benchmark, i.e., the percentage of ill-typed mutants
 over the number of mutants, ranges between 53\% (\texttt{tetris})
 and 95\% (\texttt{kCFA}). Finally, the figure depicts the individual success rate
 of mutators per benchmark which reveals that there is at
 least one benchmark where each mutator performs well
 (>90\%).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/code-hit-ratios}
  \caption{Success rate of mutators (the percentage of ill-typed mutants
  over the number of mutants)}
  \label{fig:mutators-static-hit}
\end{figure}

%% ll: Is this still necessary?
As a
final remark in this section, with the data from
figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
we target, between 25\% and
90\% of the ill-typed mutants each mutator generates give rise to at least one interesting debugging scenario.

\subsection{What are interesting debugging scenarios?}
So far we have established that our mutators and~\citet{gtnffvf-jfp-2019}'s
benchmarks are sufficient to create a large number of realistic ill-typed Typed Racket
programs. However, in an ill-typed program with correct types, like
in our case, it is trivial to figure out which component contains the bug.
It is the component that doesn't type check. Thus the ill-typed mutants
do not offer an interesting setting to try out the debugging abilities of
rational programmers. That said, we can derive interesting debugging
scenarios by removing type specifications from some of the components of
the benchmarks. The resulting mix-typed programs may be well-typed and when
run may raise a type error. Section~\ref{sec:rational} explains how we can
explore the space of debugging scenarios in a principled manner starting from
an interesting scenario.
In the rest of this section we describe how we select interesting scenarios.
We break this decision into three basic factors, each of which specify an element defining our selection of interesting scenarios.

First, not all debugging scenarios are suitable for evaluating blame across all three semantics.
For instance, there exist scenarios where Natural's checks discover a problem but Transient and Erasure do not.
Since Transient and Erasure provide no debugging information (and in particular no blame) in such cases, we have nothing to evaluate.
As alluded to in the previous paragraph, scenarios where the type checker detects the problem statically are also uninteresting.
Hence, we select only scenarios which are interesting in all three semantics.
Based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in erasure implies a dynamic type error in both Transient and Natural, for any scenario that Erasure detects the problem, both Transient and Natural should detect it as well.
We can therefore define scenarios that raise a dynamic type error in Erasure as interesting, and all such scenarios should raise dynamic type errors in Transient and Natural, too.

The second factor is that we would like to be able to compare the behavior of blame between each of our three semantics.
In order to do so, we must evaluate each of the semantics on the same set of scenarios;
this allows us to compare for any given scenario how Natural compares to Erasure, for instance.
Hence, we pre-select a sample of scenarios which serve as the starting points to evaluate all three semantics.

Finally, the third factor is that we only want to consider debugging scenarios in which the bug has effects that cross component boundaries.
In the context of migratory typing, dynamic type enforcement occurs when values cross component boundaries.
However, a bug may be caught by the runtime before any values it influences flow out of the containing component -- for instance, if the bug is that a direct argument to \texttt{+} becomes a string.
We consider such scenarios trivial, becase typing the component in which the problem manifests immediately identifies that same component as the buggy one.
In an effort to avoid these trivial scenarios, we further restrict our definition of interesting debugging scenarios to be those in which the stack trace from Erasure identifies a component other than the buggy one.

Thus we have defined the set of interesting debugging scenarios as those for which Erasure raises a dynamic type error in some component other than the buggy one.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/code-dynamic-error}
  \caption{}
  \label{fig:mutators-runtime-hit}
\end{figure}

>>>>

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow
~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
A mutation is a local syntactic change of the code of a program that may
produce a bug.
\ll{Need citations for mutation}
Usually, the outcome of mutating a program
once, dubbed a mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming errors, and mutation frameworks typically come with standard mutators carefully designed to create such mutants.

However, Typed Racket's gradual type system detects shallow errors, unlike those introduced by standard mutators.
Thus, we need different mutators to evaluate the rational programmer, ones which are fine-tuned to generate shallow errors.
The table in figure~\ref{table:mutation-ops} summarizes our mutators.
For each one, the table provides a short description and an example
mutation it performs. Each mutator is inspired by the more-than-a-decade
long experience of the authors making type-level mistakes in Typed Racket.
We find that these mistakes often take non-trivial effort to debug.

@@

\subsection{Are these mutators good enough?}
\begin{itemize}
\item do these mutators work?
\begin{itemize}
\item now we need to see if these mutators actually produce interesting errors
\begin{itemize}
\item define interesting errors: type error + dynamic err w/ 3 modules on stack in a mixed configuration
\item this section will show if they do this on our benchmarks
\end{itemize}
\item do they actually produce interesting errors? -- for this need plot that's inverted version dyn-err-heatmap + total of mutants per benchmark
\item do we need all of the mutators?
\begin{itemize}
\item each one produces type errors + interesting scenarios
\item we conjecture they represent different kinds of mistakes
\end{itemize}
\end{itemize}
\end{itemize}

Of course, mutators that introduce these kinds of shallow errors directly into the benchmarks do not produce bugs that are interesting from the perspective of the rational programmer.
The benchmarks are fully typed, so Typed Racket's type checker locates them immediately.
For that reason, we apply the mutators to versions of the benchmarks with some components stripped of type annotations (dubbed \emph{configurations}).
Without the annotations, the type checker may not detect the errors in these mutant configurations.
Instead, Typed Racket's gradual type system may detect the errors dynamically.
Mutant configurations that result in dynamic errors are the ideal ground to exercise the rational programmer's ability to locate errors.

Specifically, we define an \emph{interesting debugging scenario} as a mutant configuration that
\begin{enumerate}
\item would be rejected by the type checker if we restored the missing type annotations,
\item produces a dynamic error under Erasure as-is, and
\item the stacktrace of the error mentions at least three distinct components.
\end{enumerate}

The second and third criteria require a little more explanation.
The second criterion requires that the error is detectable under Erasure to ensure that it is detectable by all three semantics.
This is a direct consequence of~\citet{gfd-oopsla-2019}'s theorem that a dynamic type error under Erasure implies a dynamic type error under both Transient and Natural.
It is worth noting that this criterion excludes from consideration all scenarios where Natural or Transient can detect the bug but Erasure cannot, and thus it favors Erasure.
However, our goal here is to evaluate blame across these three semantics, and thus we focus on systems where all of them exhibit blame.

The third criterion aims to capture debugging scenarios with a specific kind of dynamic error: one produced by the interactions between typed and untyped components.
In those cases, the rational programmer may need to examine several components to locate the source of the bad interaction.
In technical terms, we conservatively approximate these scenarios by analyzing the stacktrace of the error.
If the stack trace mentions at least two components, then this is evidence of interaction between components.
We require the third component because our benchmarks come with a driver component which is present in every stacktrace.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/mutant-breakdown}
  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
  \label{fig:mutant-breakdown}
\end{figure}

With this definition in hand, we analyze how effectively our mutators produce interesting debugging scenarios from our benchmarks.
Figure~\ref{fig:mutant-breakdown} shows the number of distinct interesting debugging scenarios produced by our mutators for each benchmark.
The columns correspond to our benchmarks, and each rows show the count of interesting debugging scenarios produced by one mutator from figure~\ref{table:mutation-ops}.
The counts are cut of at 50, so any bar reaching the right side of the plot means 50 or more scenarios.

Figure~\ref{fig:mutant-breakdown} illustrates that our mutators do indeed produce a large number of interesting debugging scenarios for all of our benchmarks.
All of the mutators together produce at least 40 interesting scenarios for every benchmark, and those scenarios are produced by at least four different mutators for each benchmark.
Thus our mutators produce a large and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one benchmark.
Some mutators clearly only target one or two of them, though, and in particular the class-focused mutators are primarily effective in \texttt{take5}.
This reflects that among our benchmarks \texttt{take5} makes the most extensive use of object-oriented features, while the other features targeted by the rest of the mutators are used more broadly.
